---
title: "Starting Smart: Improving Wordle Efficiency"
subtitle: "DSCI-250 - Final Project"
author: "Nate Segura"
date: last-modified
date-format: long
format: 
  html:
    toc: true
    number-sections: true
    self-contained: true
    df-print: paged
---

```{r}
#| message: false
#| warning: false
# Load ALL Packages

library(tidyverse)
library(lubridate)
library(knitr)
library(ggrepel)
library(patchwork)
```

# Purpose

The purpose of this project is to explore trends in the daily Wordle puzzles by the New York Times. Wordle is a word-guessing game where players have six attempts at guessing the daily five-letter word. Each guessed word returns letters highlighted in green, yellow, or grey to indicate information about the answer word. This project analyzes the usage of different words as your first Wordle guess by quantifying and comparing the returned information in terms of "Green" or "Yellow" letters.  

The goal of this project is to explore the intracacies of the popular puzzle game and hopefully provide useful information for daily players. Accompanying this report is an interactive [Shiny Web App](https://nsegura.shinyapps.io/wordle-project-shinyapp/). This app builds on the project's goal by providing a simple way to analyze your own guess words one-by-one using the plots and rankings created through this project.

# Data

The primary dataset in this project comes from the online puzzle tool "WordFinder" published by "YourDictionary", a brand owned by the digital media company "LoveToKnow Media". This dataset can be found in a table format on the tool's updated answer page: [Archive of Past Wordle Answers](https://wordfinder.yourdictionary.com/wordle/answers/). Due to the daily updates on this site, the utilized dataset ends on April 8th of 2026, when the data was most recently scraped.

To obtain this data I used webscraping with rvest to extract tables from the webpage. This step of the process was more complex than I had expected. I ran into an issue where the column headers were different than the website showed, and the "date" column did not contain the year for any of the entries. After doing some research I came to understand that the webpage was being dynamically generated, which requires the usage of rvest's `read_html_live()` function. I intially had used `read_html()`, which gathered the static tables instead. After changing this function my data finally reflected the webpage and was ready for further wrangling. 

This wrangling included binding each dataframe using the `bind_rows()` function, since each table represented one month of Wordle puzzles. After binding, each row represented a different daily puzzle. The dataset has three columns: Date, Puzzle Number, and Puzzle Answer. There were 1,755 rows, representing almost five years of Wordle puzzles

My secondary dataset in this project comes from a GitHub user who compiled over 3,000 five-letter english words. This dataset was used in order to rank the effectiveness of each word against other words. Due to the nature of Wordle's algorithm, I decided this list was the best choice. There are around 12-15,000 possible guesses in wordle, but only \~2200 potential answers. As a result, creating a ranking chart is difficult, since including all possible guesses would make an extremely long list and lose value, but including only potential answers would exclude many possible guesses. To solve this problem, I used this list that contains more words than can be answers, such as some plural and profane words, while not providing too many options such as the 15,000 word lists. The used dataset can be found [here](https://gist.github.com/daemondevin/df09befaf533c380743bc2c378863f0c).

# Data Dictionary

- `puzzle_num`: Chronological Identifier for Each Puzzle (1 = day one, 300 = day three hundred)
- `answer`: Wordle Answer for Day 
- `date`: Date of Puzzle, further split into `year`, `month`, & `day`
- `greens`: Number of Green Letters returned by guess word
- `yellows`: Number of Yellow Letters returned by guess word

# Data Processing

## Wrangling Primary Dataset

The following code fetches the primary dataset of Wordle answers and wrangles the data. Afterwards, the dataset is written as a .csv file for the accompanying Shiny Web App.

```{r}
#| message: false

# Fetch Data
answers <- read_csv("data/wordle_answers_raw.csv")

# Wrangle Data
answers1 <- answers |> 
  rename(puzzle_num = "Wordle Puzzle",    # Change Column Names
         answer = "Wordle Answer") |> 
  filter(puzzle_num != 1755) |>         # Remove Latest Answer (formatted differently)
  mutate(year = year(mdy(Date)),
         month = month(mdy(Date)),    # Seperate Date into ymd columns for time plots
         day = day(mdy(Date)))
```

```{r}
#| eval: false

# Write answers1 as a .csv for the Shiny App
# Eval set to false for rendering

write_csv(answers1, "data/wordle_answers_clean.csv")
```

## Functions for Checking

The following tabset contains a set of functions that work to quantify how helpful a guess word is. When working through this project, I asked the following main questions: 

- "On how many days would a word give me X amount of green letters"
- "On how many days would a word give me X amount of yellow letters"
- "How does this word compare to other words?"

To begin this process, I created a set of functions that output the number of green and yellow letters a given word will obtain each day. These functions are shown in the following tabset, labeled with their purpose. By default, the final function will be open for viewing. This function outputs the number of both greens and yellows for each word in a dataframe format that is ready for plotting.

::: panel-tabset

#### Greens

This function checks how many Green letters you will receive.

Formula: Check if Letter1 = Letter1, output 1 or 0 to represent yes or no. After doing this for each letter, it adds up the 1s and creates a column for greens

```{r}
check_green <- function(df, word) {
  
  df |> 
    mutate(g1 = if_else(str_sub(answer, 1, 1) == str_sub(word, 1, 1), 1, 0),
           g2 = if_else(str_sub(answer, 2, 2) == str_sub(word, 2, 2), 1, 0),
           g3 = if_else(str_sub(answer, 3, 3) == str_sub(word, 3, 3), 1, 0),
           g4 = if_else(str_sub(answer, 4, 4) == str_sub(word, 4, 4), 1, 0),
           g5 = if_else(str_sub(answer, 5, 5) == str_sub(word, 5, 5), 1, 0)) |> 
    mutate(greens = rowSums(across(7:11)))
  
}

# Example (first 10 rows)
kable(check_green(answers1, "SEVEN")[1:10,])
```

#### Yellows (Forumla)

This function creates a formula for yellows. `str_split` and `paste0` work to change the word into regular expressions.

example: - SEVEN becomes a vector of S{1} E{1} V{1} E{1} N{1}. - `str_remove` then inputs these one by one and outputs the letters remaining afterwards.

```{r}
yellow_formula <- function(answer, word) {
  
  p_letters <- paste0(unlist(str_split(word, "")), "{1}")
  
  remaining <- 
    str_remove(
      str_remove(
        str_remove(
          str_remove(
            str_remove(
              answer, p_letters[1]), p_letters[2]), p_letters[3]), p_letters[4]), p_letters[5])
  
}
```

#### Yellows & Greens {.active}

This function builds on the previous by counting the remaining letters and subtracting them from 5. This will output the number of letters that the input word removed. Finally, the number of greens is subtracted from this number to output the number of yellows

```{r}
check_yellowgreen <- function(df, word) {
 
  stopifnot(is.data.frame(df))      # checks that df is a dataframe
  stopifnot(str_length(word) == 5)   # checks that word input is 5 letters
  
  df2 <- check_green(df, word)
  
  df3 <- df2 |> 
    select(-g1, -g2, -g3, -g4, -g5) |> 
    mutate(yellows = (5 - str_length(yellow_formula(df$answer, word))) - greens)
  
  return(df3)
}

# Test
check_yellowgreen(answers1, "CLEAR")
```
:::

## Functions for Plotting

The next step in this process was to create functions for plotting the number of yellow and green letters that the previous functions obtained. The following tabset also contains three functions, ordered similarly to the previous tabset. The default open tab contains the function for plotting both green and yellow letters side by side. This function is used for analysis and within the accompanying Shiny Web App.

The first two functions in this tabset create simple and unfinished bar charts, demonstrating the early versions of visualizations for this analysis. The first depicts the distrubtion of Green Letters returned, using the new `check_yellowgreen()` function. The second utilizes the same function to depict the distribution of Yellow Letters returned by a given guess word.

::: panel-tabset
#### Greens

```{r}
plot_green <- function(df, word) {
  
  df2 <- check_yellowgreen(df, word)
  
  df3 <- df2 |> 
           count(greens)
  
  df3 |> 
    filter(greens != 0) |> 
    ggplot(aes(greens, n)) +
    geom_col(fill = "#00BA42") +
    geom_text(aes(label = n), vjust = -1) +
    scale_x_continuous(breaks = c(1,2,3,4,5)) +
    labs(title = paste0("Distribution Green Letters in ", word),
         subtitle = "Wordle Answers from September 2021 to April 2026",
         caption = paste0("Instances of No Greens: ", sum(df2$greens == 0)))
  
}

# Test
plot_green(answers1, "ROUGH")
```

#### Yellows

```{r}

plot_yellow <- function(df, word) {
  
  df2 <- check_yellowgreen(df, word)
  
  df3 <- df2 |> 
           count(yellows)
  
  df3 |> 
    filter(yellows != 0) |> 
    ggplot(aes(yellows, n)) +
    geom_col(fill = "#FFDD33") +
    geom_text(aes(label = n), vjust = -1) +
    scale_x_continuous(breaks = c(1,2,3,4,5)) +
    labs(title = paste0("Distribution of Yellow Letters in ", word),
         subtitle = "Wordle Answers from September 2021 to April 2026",
         caption = paste0("Instances of No Yellows: ", sum(df2$yellows == 0)))
  
  
}

# Test
plot_yellow(answers1, "ROUGH")
```

#### Both {.active}

```{r}
plot_yellowgreen <- function(df, word) {
  
  stopifnot(is.data.frame(df))
  stopifnot(str_length(word) == 5)   # Safety Checks
  
  df2 <- check_yellowgreen(df, word)
  
  dfg <- df2 |> 
    count(greens) |> 
    complete(greens = 0:5, fill = list(n = 0)) |>   # Add Green Counts (include zeros)
    rename(count = greens) |> 
    mutate(color = "greens") |> 
    filter(count != 0)
  
  dfy <- df2 |> 
    count(yellows) |> 
    complete(yellows = 0:5, fill = list(n = 0)) |>    # Add Yellow Counts (include zeros)
    rename(count = yellows) |> 
    mutate(color = "yellows") |> 
    filter(count != 0)
  
  df3 <- full_join(dfg, dfy)
  
  df3 |> 
    ggplot(aes(count, n, fill = color)) +
    geom_col(position = "dodge") +
    scale_fill_manual(values = c("greens" = "#00BA42", "yellows" = "#FFDD33")) +
    geom_text_repel(aes(label = n), 
                    vjust = -1, 
                    ylim = c(0, NA),
                    position = position_dodge(width = 0.9)) +
    #scale_color_manual(values = c("greens" = "#00BA42", "yellows" = "#FFDD33")) +
    scale_x_continuous(breaks = c(1,2,3,4,5)) +
    ylim(0, max(df3$n) + 100) +
    labs(title = paste0("Distribution of Yellow and Green Letters in ", word),
         subtitle = "Wordle Answers from September 2021 to April 2026",
         caption = c(paste0("Instances of No Yellows: ", sum(df2$yellows == 0)),
                     paste0("Instances of No Greens: ", sum(df2$greens == 0))),
         x = "Number of Green or Yellows",
         y = "") +
    theme(legend.position = "none",
          plot.caption = element_text(hjust = c(1, 0)))
}
  
# Test
plot_yellowgreen(answers1, "ROUGH")
```
:::

## Functions for Ranking

Finally, this third tabset contains functions that aim to answer my third question: "How does this word compare to other words?" Utilizing the previous functions that were designed for outputting plottable tables, these two functions each output a simple vector containing the total number of green or yellow letters achieved by a single word. By using these functions, we can apply a numerical value to any five letter word and compare its value against others.

::: panel-tabset
#### All Greens

```{r}
all_green <- function(df, word) {
  
  df1 <- check_yellowgreen(df, word)
  
  return(sum(df1$greens))
}

# Test
all_green(answers1, "SEVEN")
```

#### All Yellows

```{r}
all_yellow <- function(df, word) {
  
  df1 <- check_yellowgreen(df, word)
  
  return(sum(df1$yellows))
}

# Test
all_yellow(answers1, "SEVEN")
```
:::

## Wrangling Secondary Dataset

This code fetches the data from my secondary dataset, which contains a list of over 3,000 five-letter words. This data is first mutated to a proper format. Next, the functions `all_green()` and `all_yellow()` are used to establish a total number of returned Green or Yellow letters for each five-letter word.

```{r}
#| cache: true
#| message: false

allwords <- read_csv("data/english_words.csv")

wordrankings <- allwords |> 
  mutate(X1 = str_to_upper(X1)) |> 
  rowwise() |> 
  mutate(totalgreen = all_green(answers1, X1),
         totalyellow = all_yellow(answers1, X1)) |> 
  ungroup()
```

Next, this data is formatted further to create better tables in visualizations.

```{r}
wordrankings2 <- wordrankings |>
  rename(guess = X1) |> 
  mutate(greenrank = paste0("#", min_rank(-totalgreen)),
         yellowrank = paste0("#", min_rank(-totalyellow)))
```

Lastly, this data is saved for usage in the Shiny App. 

```{r}
#| eval: false

# Write Ranked Frame as a .csv for Shiny App
# Eval set to false for rendering

write_csv(wordrankings2, "data/english_words_ranked.csv")
```


# Analysis

## Rankings

This table shows the 10 words with the highest total count of green letters achieved. These results mirror common reccomendations, which often suggest "slate" as the best starting word for any Wordle puzzle. We can also see other popular reccomendations such as "crane" and "snare". Interestingly, some of the top guesses for green letters perform quite poorly in yellow letters. The 4th best word for green letters, "crane", is placed 2706th for yellows, among the lowest totals for any word (3,103 total).

```{r}
wordrankings2 |> 
  slice_max(totalgreen, n = 10)
```

This next table shows the top 10 words with the highest total count of yellow letters. Similar to the green rankings from before, many guesses in this top ten list hold extremely low green rankings. The number one guess for yellow letters, "opera", is among the worst 100 words for green letters in this dataset. This disparity shown in both tables is likely due to a lack of intersections between green and yellow letters. This oversight is discussed further in the reflection portion of this project. Despite this, we can see an interesting trend of vowels in these words, with opera containing three different vowels. The inclusion of multiple vowels appears important for maximizing yellow letters, though inefficient for green letters.

```{r}
wordrankings2 |> 
  slice_max(totalyellow, n = 10)
```

## Analyzing "Slate" and "Audio"

As shown through the ranking tables, the word "slate" stands out as the alleged best guess for green letters in Wordle. This comes without suprise, as even Wordle's own "Wordle Bot" suggests using the word "slate" as a starting point. Another standoutword is the popular "audio". This word is widely popular for its use of four vowels, helping to narrow down words by attacking a vital structural mechanism. As previously mentioned regarding the ranking table for yellow letters, maximizing vowels appears to be quite useful.

Using the previously created functions, we can actually assess how the word "audio" performs and how it shapes up against others such as "slate". The following plots shows the distribution of green and yellow letters for the words "slate" (left) and "audio" (right).

```{r}
#| warning: false
#| message: false

p1 <- plot_yellowgreen(answers1, "SLATE") +
  labs(title = "Distribution of Yellow and Green Letters in SLATE (left) and AUDIO (right)",
       caption = "Instances of No Yellows: 547\nInstances of No Greens: 916") +
  theme(plot.caption = element_text(hjust = 0))

p2 <- plot_yellowgreen(answers1, "AUDIO") +
  labs(title = "",
       subtitle = "",
       caption = "Instances of No Yellows: 297\nInstances of No Greens: 1333") +
  theme(plot.caption = element_text(hjust = 1))

p1 + p2 + 
  plot_layout(axis_titles = "collect") & ylim(0, 1170)
```

These plots present the large gap between these two words, apparent in the leftmost category, instances of one produced green or yellow letter. While slate performs seemingly even in green and yellow letters, audio produces a much higher instance of yellow letters. This lines up with its usage for assessing vowels by showing the consistent usefulness of the word, even if actual placement is not discovered.

Additionally, we can compare the two words rankings. The following table shows the rankings of both "audio" and "slate".

```{r}
wordrankings2 |> 
    filter(guess == "SLATE" | guess == "AUDIO")
```
Suprisingly, both are quite similar in their yellow rank, with "slate" just barely surpassing "audio" with only 10 more achieved yellow letters, placing the word 18 ranks higher. At first this seems shocking, since the previous plots depict audio as a guess that produces a huge amount of yellow letters, however, slate outperforms audio in the two-letter and three-letter categories, both of which hold higher weight for counting overall yellow letters achieved. Through this we can conclude that despite audio's clear ability to assess vowel usage, its performance does not hold up against other guesses.

## Over Time

One final way to analyze wordle guesses can be to see their effectiveness over time. The following plot shows the average amount of yellow and green letters achieved by the top ranked word "slate" by month in 2025. This plot depicts a few intersting subtle trends for the #1 guess word. Most apparent is the spike in green letters during June, accompanied by a dip in yellow letters. This spike and dip is followed by an inverse effect in July, where green letters reaches its lowest average and yellow letters spikes back up.

```{r}
slateframe <- check_yellowgreen(answers1, "SLATE") |> 
  filter(year == 2025) |> 
  group_by(month) |> 
  mutate(g_avg = mean(greens),
         y_avg = mean(yellows))

slateframe |> 
  ggplot(aes(x = month)) +
  geom_line(aes(y = y_avg), color = "#FFDD33", linewidth = 0.8) +
  geom_point(aes(y = y_avg), color = "#FFDD33") +
  geom_line(aes(y = g_avg), color = "#00BA42", linewidth = 0.8) +
  geom_point(aes(y = g_avg), color = "#00BA42") +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  labs(title = "Effectiveness of Slate by Month in 2025",
       subtitle = "Values are Month Averages of Green and Yellow Letters Achieved",
       x = "Month",
       y = "Average Letters Achieved")
```
## The Shiny App

The final piece of analysis for this project is an accompanying Shiny Web App. This app uses Posit's "shinyapps.io" to create a standalone functional app that provides information for any inputted word. The code for this app can be found in "03_final-draft_shiny.R". The app returns a table with your word's rankings and a bar chart depicting green and yellow letters. This bar chart is the third and primary visualization created in this project's "Functions for Plotting" section. The app also includes a summarized description of the project to allow for independent usage and understanding of the app.

The app can be used at the following link: <https://nsegura.shinyapps.io/wordle-project-shinyapp/>

# Reflection

Overall I found this project to be a fun analysis and I'm proud to have created a functional shiny web app. That being said, I believe there are a few points for improvement. If I had more time on this project I would likely attempt to reduce the amount of functions needed for my plots and analysis. In my process I simply used the first functional iterations of my ideas rather than optimizing for reduced code. Additionally, I would like to further use my time analysis and implement it into my Shiny web app. Currently this analysis is a small portion of the project and only used on the word "slate".

Lastly, an issue with my method of analysis regards the intersection between Green and Yellow letters produced. As previously mentioned in my rankings analysis, this project lacks an analysis that combines Green and Yellow letters. In attempting this, I found it difficult to make a concrete ranking using combinations of both categories. In further research I found more advanced measures that can assess a word's efficiency. If I was able to expand on this project I would dive into these measures, specifically the idea of partitioning, which examines how a guess word reduces the amount of possible answers. One article I found that dives deeper into this is ["Guessing with More Skill" by Josh Stephenson](https://medium.com/@joshstephenson/the-economics-of-guessing-7c00062c2370). With more time and resources, I would like to analyze Wordle further using similar methods. I think these methods would greatly enhance the ability of my Shiny App in providing an analysis for any given word.






